---
title: "Mapping Urban and Suburban Space in the Houston MSA"
author: "Samane Afshari"
format: pdf
editor: visual

execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(tidycensus)
library(tigris)
library(tidyverse)
library(sf)
library(ggplot2)
library(scales)

options(tigris_use_cache = TRUE)


```

# Introduction

Houston was selected for this project because it is one of the largest and fastest-growing metropolitan areas in the United States and represents a highly decentralized form of urban development. Unlike older monocentric cities that are organized around a single dense downtown core, Houston exhibits extensive suburban expansion and multiple activity centers connected through highways and automobile-oriented infrastructure. This makes Houston an appropriate case for examining the spatial distribution of urban, suburban, and exurban landscapes through Census tract population density classifications.

The landscape map demonstrates that the highest-density urban tracts are concentrated near central Houston and surrounding inner-city neighborhoods. Urban and suburban tracts extend outward along major transportation corridors, while exurban landscapes dominate the outer metropolitan fringe. The overall pattern reflects Houston’s large spatial footprint, continued suburban expansion, and relatively low-density metropolitan growth compared to older northeastern cities.

```{r}
# census_api_key("51b09e0aa87b1b3a4521e4b1cb12e2ae4f8b1a61", install = TRUE)

# 1. Houston MSA boundary
cbsa <- core_based_statistical_areas(year = 2020, cb = TRUE)

houston_msa <- cbsa %>%
  filter(str_detect(NAME, "Houston")) %>%
  st_transform(3083)

# 2. 2020 Census tract population
tract_pop <- get_decennial(
  geography = "tract",
  variables = "P1_001N",
  state = "TX",
  year = 2020,
  sumfile = "pl",
  geometry = TRUE
) %>%
  st_transform(3083)

# 3. Keep tracts inside Houston MSA
houston_tracts <- tract_pop %>%
  st_filter(houston_msa, .predicate = st_intersects) %>%
  rename(population = value)

# 4. Calculate area and density
houston_tracts <- houston_tracts %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1000000,
    density_km2 = population / area_km2,
    landscape = case_when(
      density_km2 < 250 ~ "Exurban",
      density_km2 >= 250 & density_km2 < 550 ~ "Suburban low",
      density_km2 >= 550 & density_km2 < 800 ~ "Suburban high",
      density_km2 >= 800 & density_km2 < 1900 ~ "Urban low",
      density_km2 >= 1900 ~ "Urban high"
    ),
    landscape = factor(
      landscape,
      levels = c("Exurban", "Suburban low", "Suburban high", "Urban low", "Urban high")
    )
  )

# 5. Main landscape map
landscape_map <- ggplot(houston_tracts) +
  geom_sf(aes(fill = landscape), color = NA) +
  theme_void() +
  labs(
    title = "Population-density landscapes in the Houston MSA",
    subtitle = "2020 Census tracts classified using Hanberry-style density thresholds",
    fill = "Landscape"
  )


ggsave("figures/houston_landscape_map.png", landscape_map, width = 9, height = 7, dpi = 300)
print(landscape_map)
```

```{r}
# ACS
acs_vars <- c(
  median_income = "B19013_001",
  total_poverty = "B17001_001",
  poverty_count = "B17001_002"
)

acs_2020 <- get_acs(
  geography = "tract",
  variables = acs_vars,
  state = "TX",
  year = 2020,
  survey = "acs5",
  geometry = FALSE
) %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(poverty_rate = poverty_count / total_poverty * 100)

houston_acs <- houston_tracts %>%
  left_join(acs_2020, by = "GEOID")

# map 2
income_map <- ggplot(houston_acs) +
  geom_sf(aes(fill = median_income), color = NA) +
  scale_fill_viridis_c(labels = dollar) +
  theme_void() +
  labs(
    title = "Median household income by Census tract",
    subtitle = "Houston MSA, 2016–2020 ACS 5-year estimates",
    fill = "Income"
  )

ggsave("figures/houston_income_map.png", income_map, width = 9, height = 7, dpi = 300)
print(income_map)
```



```{r}
library(scales)
library(forcats)

income_chart <- houston_acs %>%
  st_drop_geometry() %>%
  group_by(landscape) %>%
  summarize(
    median_income = median(median_income, na.rm = TRUE)
  ) %>%
  
  # reorder categories
  mutate(
    landscape = fct_reorder(landscape, median_income)
  ) %>%
  
  ggplot(aes(x = landscape, y = median_income)) +
  
  geom_col(width = 0.7, fill = "steelblue") +
  
  # value labels
  geom_text(
    aes(label = dollar(median_income)),
    vjust = -0.4,
    size = 4
  ) +
  
  scale_y_continuous(
    labels = dollar,
    expand = expansion(mult = c(0, 0.08))
  ) +
  
  labs(
    title = "Median Household Income",
    subtitle = "Houston metropolitan area",
    x = NULL,
    y = "Median household income"
  ) +
  
  theme_minimal(base_size = 13) +
  
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )


print(income_chart)

ggsave(
  "figures/income_by_landscape.png",
  income_chart,
  width = 8,
  height = 5,
  dpi = 300
)
```




```{r}
# population pyrimad 
age_vars <- load_variables(2020, "acs5", cache = TRUE) %>%
  filter(str_detect(name, "^B01001_"))

age_data <- get_acs(
  geography = "tract",
  table = "B01001",
  state = "TX",
  year = 2020,
  survey = "acs5",
  geometry = FALSE
)

# variables into sex + age groups.

# Download age-sex data
age_data <- get_acs(
  geography = "tract",
  table = "B01001",
  state = "TX",
  year = 2020,
  survey = "acs5",
  geometry = FALSE
)

# Join with your Houston landscape categories
age_data <- age_data %>%
  left_join(
    houston_tracts %>%
      st_drop_geometry() %>%
      select(GEOID, landscape),
    by = "GEOID"
  )


# recode age group

  library(dplyr)
library(stringr)

age_pyramid <- age_data %>%
  mutate(
    
    sex = case_when(
      variable >= "B01001_003" & variable <= "B01001_025" ~ "Male",
      variable >= "B01001_027" & variable <= "B01001_049" ~ "Female"
    ),
    
    age_group = case_when(
      variable %in% c("B01001_003", "B01001_027") ~ "0-4",
      variable %in% c("B01001_004", "B01001_028") ~ "5-9",
      variable %in% c("B01001_005", "B01001_029") ~ "10-14",
      variable %in% c("B01001_006", "B01001_030") ~ "15-17",
      variable %in% c("B01001_007", "B01001_031") ~ "18-19",
      variable %in% c("B01001_008", "B01001_009",
                      "B01001_032", "B01001_033") ~ "20-24",
      variable %in% c("B01001_010", "B01001_011",
                      "B01001_034", "B01001_035") ~ "25-34",
      variable %in% c("B01001_012", "B01001_013",
                      "B01001_036", "B01001_037") ~ "35-44",
      variable %in% c("B01001_014", "B01001_015",
                      "B01001_038", "B01001_039") ~ "45-54",
      variable %in% c("B01001_016", "B01001_017",
                      "B01001_040", "B01001_041") ~ "55-64",
      variable %in% c("B01001_018", "B01001_019",
                      "B01001_020", "B01001_021",
                      "B01001_042", "B01001_043",
                      "B01001_044", "B01001_045") ~ "65-74",
      variable %in% c("B01001_022", "B01001_023",
                      "B01001_046", "B01001_047") ~ "75-84",
      variable %in% c("B01001_024", "B01001_025",
                      "B01001_048", "B01001_049") ~ "85+"
    )
  ) %>%
  filter(!is.na(age_group))
```



```{r}
# pyriamd 
urban_summary <- age_pyramid %>%
  group_by(sex, age_group) %>%
  summarize(
    population = sum(estimate, na.rm = TRUE),
    .groups = "drop"
  )


suburban_summary <- age_pyramid %>%
  group_by(sex, age_group) %>%
  summarize(population = sum(estimate, na.rm = TRUE))

# step 6
urban_summary <- urban_summary %>%
  mutate(
    population = ifelse(sex == "Male", -population, population)
  )

suburban_summary <- suburban_summary %>%
  mutate(
    population = ifelse(sex == "Male", -population, population)
  )

# step 7
# urban plot
urban_plot <- ggplot(
  urban_summary,
  aes(x = population, y = age_group, fill = sex)
) +
  geom_col() +
  scale_x_continuous(labels = abs) +
  theme_minimal() +
  labs(
    title = "Urban population pyramid",
    x = "Population",
    y = "Age group"
  )
print(urban_plot)
```




```{r}
# suburban plot
suburban_plot <- ggplot(
  suburban_summary,
  aes(x = population, y = age_group, fill = sex)
) +
  geom_col() +
  scale_x_continuous(labels = abs) +
  theme_minimal() +
  labs(
    title = "Suburban population pyramid",
    x = "Population",
    y = "Age group"
  )
print(suburban_plot)
```



#  Interpretation
The figures reveal clear demographic and socioeconomic differences across Houston’s urban and suburban landscapes. The population density map shows a strong concentration of dense urban tracts near the metropolitan core, while lower-density suburban and exurban areas dominate the outer portions of the region. The median household income chart indicates that suburban landscapes, particularly suburban low-density areas, have the highest household incomes, while urban high-density areas have the lowest median incomes. This pattern suggests that wealthier households are concentrated in lower-density suburban environments, whereas denser urban neighborhoods include larger shares of lower-income populations. The population pyramids also show differences between urban and suburban demographic structures. Both landscapes contain large working-age populations, but suburban areas appear to include relatively larger family-age cohorts and older adults, reflecting Houston’s suburban residential development patterns.
